---
title: "p8105_hw3_lz3014"
auther: "Liqi Zhou"
date: 2024-10-14
output: github_document
---

```{r}
library(tidyverse)
```


# Problem 1
```{r}
library(p8105.datasets)
data("ny_noaa")
```

### 1.Write a short description of the dataset.
This dataset contains `r nrow(ny_noaa)` rows and `r ncol(ny_noaa)` columns, including seven variables: `id` (weather station id), `date` (date of observation), `prcp` (total daily precipitation), `snow` (snowfall), `snwd` (snow depth), `tmax` and `tmin` (min and max temperature).

### 2.Do some data cleaning. 
```{r}
ny_noaa_clean <- ny_noaa |>
  separate(date, into = c("year", "month", "day"), convert = TRUE) |>
  mutate(tmax = as.numeric(tmax), tmin = as.numeric(tmin))
```

- For snowfall, what are the most commonly observed values? Why?
0 is the most commonly observed value for snowfall, because most days of the year, it does not snow in NY.
```{r}
ny_noaa_clean |>
  count(snow) |>
  arrange(desc(n))
```

### 3.Make a two-panel plot showing the average max temperature in January and in July in each station across years.
```{r}
ny_noaa_clean |>
  filter(month %in% c(1, 7)) |>
  group_by(id, year, month) |>
  summarise(mean_tmax = mean(tmax, na.rm = TRUE)) |>
  ggplot(aes(x = year, y = mean_tmax, color = as.factor(month))) +
  geom_point() +
  geom_line() +
  facet_wrap(~ month, scales = "free_y") +
  labs(
    title = "Average Max Temperature in January and July across Stations",
    x = "Year",
    y = "Average Max Temperature (°C)",
    color = "Month"
  )
```

- Is there any observable / interpretable structure? Any outliers?
the mean temperature in January is significantly lower than in July across all stations and years. Stations generally show similar temperature trends over time—when one station records a higher temperature for a given year, others tend to follow the same pattern. However, we observe an unusually cold station in July of 1987 or 1988, along with a few other minor outliers.

### 4.Make a two-panel plot
- (i) tmax vs tmin for the full dataset (note that a scatterplot may not be the best option)
```{r}
ny_noaa_clean |>
  ggplot(aes(x = tmin, y = tmax)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "lm", color = "lightblue", se = FALSE) + 
  labs(
    title = "Tmax vs Tmin",
    x = "Minimum Temperature (°C)",
    y = "Maximum Temperature (°C)"
  )
```

- (ii) make a plot showing the distribution of snowfall values greater than 0 and less than 100 separately by year.
```{r}
ny_noaa_clean |>
  filter(snow > 0, snow < 100) |>
  ggplot(aes(x = snow)) +
  geom_histogram(binwidth = 5, fill = "lightblue", color = "lightyellow") +
  facet_wrap(~ year) +
  labs(
    title = "Distribution of Snowfall by Year",
    x = "Snowfall",
    y = "Count"
  )
```


# Problem 2
### 1.Load, tidy, merge, and otherwise organize the data sets.
```{r}
covar <- read_csv(file = "./data/nhanes_covar.csv", skip = 4, na = c("NA",".","")) |>
  janitor::clean_names() |>
  filter(age >= 21) |>
  drop_na()

accel <- read_csv(file = "./data/nhanes_accel.csv", na = c("NA",".","")) |>
  janitor::clean_names() |>
  drop_na()
```

- merge and tidy
```{r}
mims <- covar |>
  left_join(accel, by = "seqn")

mims_tidy <- 
  pivot_longer(
    mims,
    min1:min1440,
    names_to = "mins",
    names_prefix = "min",
    values_to = "accel_min_data"
  ) |>
  mutate(mins = as.numeric(mins))
```

- encode data with reasonable variable classes
```{r}
mims_tidy <- mims_tidy |>
  mutate(sex = factor(sex, levels = c(1, 2), 
                      labels = c("Male", "Female")), 
         education = factor(education, levels = c(1, 2, 3), 
                            labels = c("Less than high school", "High school equivalent", "More than high school")))
mims_tidy
```


